---
title: "RPPA Exploratory Analysis: ASNS"
author: 'Nisha Narayanan (nnarayanan@mdanderson.org)'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: flatly
    highlight: pygments
---
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

Install the packages. Since the packages are already installed, chunk valuation is set to FALSE. Run manually if required. 
```{r pkg, warning=FALSE, message=FALSE, eval=FALSE}
install.packages("BiocManager")
install.packages(c("ggplot2", "dplyr", "readxl", "xlsx"))
install.packages("tidyr")
install.packages("ggsurvfit")
install.packages("tidycmprsk")
install.packages("ggpubr")
install.packages("gtsummary")
BiocManager::install("phenoTest")
install.packages("magrittr")
install.packages("survivalAnalysis")
```

Load the libraries required for downstream analysis
```{r lib, warning=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(ggsurvfit)
library(tidycmprsk)
library(ggpubr)
library(gtsummary)
library(survival)
library(phenoTest)
library(survminer)
library(survivalAnalysis)
```

## Data loading

Load the data into the R environment. Accessing the excel sheet through the read_excel function in the readxl library.

```{r data}
rppa_data <- read_excel("/Users/nnarayanan/Library/CloudStorage/OneDrive-InsideMDAnderson/Data/RPPA/AML_A3_Patient-n810_Protein-n429_Data.xlsx") #Replace path to file with the actual path to the file in your machine
```

## Data cleaning

This step focuses on cleaning the data for subsequent analysis. 
1. The column names of some of the columns in the RPPA data needs to be changed for some reasons (like duplicated column names), rename them if it is going to be used in downstream analyses. 
2. We also have a lot of NA data in the data frame which gets read as a character string in R, these values need to be changed to NA (missing value). Inspect the data frame and identify these columns of concerns.  
3. For most of the statistical analysis and plots we need variables that are grouped together by factors. Identify these column numbers and convert them to factors using lapply.  

```{r data_fil}
#names(rppa_data)[23:26] <- c("Mono", "Blasts_Presence", "CD34", "CD14") #updating duplicate column names

names <- c("Race.Dx", "FAB.Assigned", "Prior.Malignancy", "CNS.Leukemia", "TX1.BestResponse2", "Relapse", "WBC.hyper", "Complex.Karyotype", "MD.FLT3.ITD.mutated", "Cytogenetics.Risk.Group", "Cytogenetics.Cat.Summary", "Performance.Status", "MD.MLL_AF4.mutated", "MD.Risk.Group", "MD.CEBPA.mutated", "TX1.Remission", "PseudoFAB", "Prior.Chemo", "PseudoFAB", "MD.Risk.Group","TX1.Remission", "Performance.Status" ) # identify any or all such rows and store the column names in this variable
for(i in names){
  #print(i)
  rppa_data[[i]] <- gsub("NA", NA, rppa_data[[i]]) #replacing NA character with NA
}
rppa_data$Cytogenetics.Risk.Group <- ifelse(rppa_data$Cytogenetics.Risk.Group=="U", "Unfavorable", ifelse(rppa_data$Cytogenetics.Risk.Group=="I", "Intermediate", ifelse(rppa_data$Cytogenetics.Risk.Group=="F", "Favorable", NA)))
rppa_data[c(4, 6,7, 12:14,19:23,28,51,65,66,260, 262,376:401)] <- lapply(rppa_data[c(4,6,7, 12:14, 19:23, 28, 51, 65, 66, 260, 262,376:401)], factor) #Converting the columns of required data frame to factor variables
```

## Data Summary

This table summarizes the demographic data of patients.

```{r}
tbl_summary(
    rppa_data,
    include = c(Age.Dx, Diagnosis, Race.Dx, FAB.Assigned),
    by = Gender, # split table by group
    missing = "ifany" # don't list missing data separately
  ) %>%
  add_n() %>% # add column with total number of non-missing observations
  add_p() %>% # test for a difference between groups
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels()
```


## Boxplots 

This section focuses on constructing box plots for the selected variables against the protein of interest. We also calculate the p-values using either t-test (<=2 groups) or one-way ANOVA (>2 groups). The same p-value will be printed on the plots. 

This chunk calculates the p-value for all variables of interest

```{r}
## p-val helper function
col_names <- colnames(rppa_data)[c(4, 6,7, 12:14,19:23,28,51,65,66,260, 262,376:401)] #select all the variables you want to explore
poi <- "ASNS" #replace with the name of your protein of interest
stat_res <- data.frame(Variable = character(), n = integer(), Test = character(), P.Val = numeric()) #creating an empty data frame to store the results. 
i=1
for(var_name in col_names){
  #print(var_name)
  df <- rppa_data %>% dplyr::select(poi, var_name)
  df <- na.omit(df) #omiting NA values as it is not required for statistical analysis
  n = nrow(df)
  if(length(levels(rppa_data[[var_name]])) <=2) {
          value <- dplyr::pull(df[,poi]) #extracting BCL11A values into a vector
          var <- dplyr::pull(df[,var_name]) #extracting other variables as required
          x <- t.test(value ~ var) #t-test
          pval = x$p.value #extracts and stores the p-value in the results df
          test = "T-test" #name of the test is also stored in results df
  }
  else{
      value <- dplyr::pull(df[,poi]) #extracting BCL11A values into a vector
      var <- dplyr::pull(df[,var_name]) #extracting other variables as required
      atest <- aov(value ~ var) #anova
      summ <- summary(atest)
      pval_list = summ[[1]][["Pr(>F)"]]
      pval = pval_list[1] #extracts and stores the p-value in the results df
      test = "ANOVA" #name of the test is also stored in results df
  }
      stat_res[i,] <- c(var_name, n, test, pval)
      i=i+1
      df <- data.frame()

}
class(stat_res$P.Val) <- "numeric" 
stat_res$P.Val <- signif(stat_res$P.Val,3) #rounding the P-value to 3 significant digits
```

Helper function for all boxplots.

```{r}
# Helper function for boxplot
box_plot <- function(df, x, y) {
  #lab = df[i,3]
  #lab = round(lab, digits=4)
  ggplot(df, aes(x = !! sym(x), y = !! sym(y), fill = !! sym(x))) + geom_boxplot() + labs(title = paste0(y, " expression across ", x, " groups"), caption = paste0(stat_res$Test[i], ": p-value = ", stat_res$P.Val[i], "\n", "n=", stat_res$n[i])) +   
    geom_text(data=df %>%
              group_by(!! sym(x)) %>% 
              dplyr::summarise(top = max(!! sym(y)), n=dplyr::n()), 
            aes(x=!! sym(x), y=max(rppa_data[[poi]]), label= paste0("n = ", n)), 
            nudge_y=1) + 
    theme(panel.background=element_rect(fill="white"), axis.line.x.bottom=element_line(color="black"), axis.line.y.left=element_line(color="black"))
}
```

Main loop through the boxplots
```{r}
# Main loop through the columns and dataset
i=1
for(var in col_names){
  print(box_plot(df=rppa_data, x=var, y=poi))
  i=i+1
}
```

Use the below chunk if your plot needs to be altered in some ways to make it more readable. For instance, our RPPA data has a Cytogenetics.Cat.Summary column which when looked with BCL11A, had too many row names making it hard to interpret. This chunk rotates the axis labels, and adjusts the font size and position to fix this. Corresponding statistical variables where also extracted. Likewise, many other plot related issues can be fixed by altering some of the parameters without having to change the other plots.

```{r, fig.width=10}
ggplot(rppa_data, aes(x = Cytogenetics.Cat.Summary, y = ASNS, fill = Cytogenetics.Cat.Summary)) + geom_boxplot() + labs(title = paste0("ASNS expression across cytogenetic cat summary groups"), caption = paste0(stat_res$Test[17], ": p-value = ", stat_res$P.Val[17], "\n", "n=", stat_res$n[17])) +   
    geom_text(data=rppa_data %>%
              group_by(Cytogenetics.Cat.Summary) %>% 
              dplyr::summarise(top = max(ASNS), n=dplyr::n()), 
            aes(x=Cytogenetics.Cat.Summary, y=max(rppa_data[[poi]]), label= paste0("n = ", n)), 
            nudge_y=1) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + theme(panel.background=element_rect(fill="white"), axis.line.x.bottom=element_line(color="black"), axis.line.y.left=element_line(color="black"))

ggplot(rppa_data, aes(x = PseudoFAB, y = ASNS, fill = PseudoFAB)) + geom_boxplot() + labs(title = paste0("ASNS expression across Pseudo FAB groups"), caption = paste0(stat_res$Test[3], ": p-value = ", stat_res$P.Val[3], "\n", "n=", stat_res$n[3])) +   
    geom_text(data=rppa_data %>%
              group_by(PseudoFAB) %>% 
              dplyr::summarise(top = max(ASNS), n=dplyr::n()), 
            aes(x=PseudoFAB, y=max(rppa_data[[poi]]), label= paste0("n = ", n)), 
            nudge_y=1) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + theme(panel.background=element_rect(fill="white"), axis.line.x.bottom=element_line(color="black"), axis.line.y.left=element_line(color="black"))
```

## Scatter plots

```{r, warning=FALSE}
# Helper function for boxplot
scatter_plot <- function(df, x, y) {
  ggplot(df, aes(x = !! sym(x), y = !! sym(y))) + geom_point() + labs(title = paste0("Pearson Correlation: ", y, " - ", x)) +   
    stat_smooth(method = "lm",
                formula = y~x,
                geom = "smooth") +
    stat_cor(method="pearson") + 
    theme(panel.background=element_rect(fill="white"), axis.line.x.bottom=element_line(color="black"), axis.line.y.left=element_line(color="black"))
}

```

```{r, warning = FALSE}
rppa_data[,c(543,546:566, 568:599)] <- sapply(rppa_data[,c(543,546:566, 568:599)],as.numeric) #convert all the variables of interest to numeric
# scatter plot call
i=1
scat_col_names <- colnames(rppa_data)[c(543,546:566, 567:599)]
for(var in scat_col_names){
  print(scatter_plot(df=rppa_data, x=var, y=poi))
  i=i+1
}
```

## Correlation Plots

Pearson's correlation is performed to study the correlation between the protein of interest and all the other proteins. You can either calculate the correlations of all proteins with all proteins or just focus on your protein of interest. Both codes are available, use the one that is needed. The write.table function also helps to export the results.

```{r}
corr_matrix <- as.matrix(rppa_data[,609:1037]) #extract the protein data into a matrix
#corr_matrix <- as.matrix(corr_matrix) 

#To look at correlation of all proteins with all proteins 
#corr_values <- as.data.frame(cor(corr_matrix, method = "pearson"))
#write.table(corr_values, "All_Protein_Correlation_Values.csv", sep = "\t", row.names = T)
#Extract protein of interest from the correlation matrix and proceed for further analysis.
#corr_values_poi <- as.data.frame(corr_values[[poi]])
#rownames(corr_values_poi) <- rownames(corr_values)
#corr_values_poi$Protein <- rownames(corr_values_poi)
#corr_values_poi <- corr_values_poi %>% arrange(corr_values[[poi]]) #replace with the actual column name

#Perform the correlation analysis of the protein of interest directly
corr_values <- as.data.frame(cor(corr_matrix[,-26], rppa_data[[poi]])) #x is the column number of your protein of interest in the corr_matrix.
names(corr_values) <- "Correlation_Values"
corr_values$Protein <- rownames(corr_values)
corr_values <- corr_values %>% arrange(Correlation_Values)
write.table(corr_values, "Protein_Correlation_Values.csv", sep = "\t", row.names = T)
```

This step does the waterfall plot for the most correlated positive and negative proteins with the protein of interest. 

```{r, fig.height, 12, fig.width=16}

pos <- corr_values %>% filter(Correlation_Values>=0) #extract positive correlation values
neg <- corr_values %>% filter(Correlation_Values<=0) #extract negative correlation values
pos <- tail(pos,100) #keeps only the top 100 positive correlation proteins
neg <- head(neg,100) #keeps only the top 100 negative correlation proteins

ggbarplot(pos, x = "Protein", y = "Correlation_Values", fill = "blue") +   
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + labs(title = "Top 100 Positively Correlated Proteins with ASNS")

ggbarplot(neg, x = "Protein", y = "Correlation_Values", fill = "red") +   
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + scale_x_discrete(position = "top") + labs(title = "Top 100 Negatively Correlated Proteins with ASNS")
```

## Survival Plots

```{r, fig.height=7, fig.width=10}
rppa_data$status <- ifelse(rppa_data$vital.status=="D", 1, 0) #including a factor column for alive and dead
rppa_data$vital.status <- factor(rppa_data$vital.status)
#cumulative incidence curve
cuminc(Surv(OS.Wks.SMK, vital.status) ~ 1, rppa_data) %>%
  ggcuminc(outcome = "D") +
  add_risktable()

rppa_data$poi_median <- ifelse(rppa_data[[poi]]>=median(rppa_data[[poi]]), "Protein Expression > Median", "Protein Expression < Median") # dividing expression by median
rppa_data$poi_median <- factor(rppa_data$poi_median)

poi_quartile <- quantile(rppa_data[[poi]],probs=c(0.25,0.5,0.75,1))
poi_tertile <- quantile(rppa_data[[poi]],probs=c(0.333,0.667,1))

rppa_data$poi_quartile<- ifelse(rppa_data[[poi]] <= poi_quartile[1], "Q1",ifelse(poi_quartile[1] < rppa_data[[poi]] & rppa_data[[poi]] <= poi_quartile[2], "Q2", ifelse(poi_quartile[2] < rppa_data[[poi]] & rppa_data[[poi]]<= poi_quartile[3],"Q3", "Q4")))

rppa_data$poi_tertile<- ifelse(rppa_data[[poi]] <= poi_tertile[1], "Q1",ifelse(poi_tertile[1] < rppa_data[[poi]] & rppa_data[[poi]] <= poi_tertile[2], "Q2", "Q3"))

rppa_data$poi_quartile <- factor(rppa_data$poi_quartile)
rppa_data$poi_tertile  <- factor(rppa_data$poi_tertile)

median_fit <- survfit(Surv(OS.Wks.SMK, status) ~ poi_median, data = rppa_data)
cox_median <- coxph(Surv(OS.Wks.SMK, status) ~ poi_median, data = rppa_data)
ggsurvplot(median_fit, data=rppa_data, pval=T, palette = c("blue", "red"),
           legend.labs = c("Low Expression", "High Expression"), xlab= "Survival Time (Weeks)",
           pval.coord = c(50,0.05), legend=c(0.75,0.9), legend.title="", title="Survival Probability (ASNS Median expression groups)", conf.int = FALSE,
           risk.table = TRUE) 

tertile_fit <- survfit(Surv(OS.Wks.SMK, status) ~ poi_tertile, data = rppa_data)
cox_tertile <- coxph(Surv(OS.Wks.SMK, status) ~ poi_tertile, data = rppa_data)
ggsurvplot(tertile_fit, data=rppa_data, pval=T, palette = c("blue", "red", "orange"),
           legend.labs = c("Q1", "Q2", "Q3"), xlab= "Survival Time (Weeks)",
           pval.coord = c(50,0.05), legend=c(0.75,0.9), legend.title="", title="Survival Probability (ASNS Tertile expression groups)", conf.int = FALSE,
           risk.table = TRUE) 

quartile_fit <- survfit(Surv(OS.Wks.SMK, status) ~ poi_quartile, data = rppa_data)
cox_quartile <- coxph(Surv(OS.Wks.SMK, status) ~ poi_quartile, data = rppa_data)
ggsurvplot(quartile_fit, data=rppa_data, pval=T, palette = c("blue", "red", "orange","magenta"),
           legend.labs = c("Q1", "Q2", "Q3", "Q4"), xlab= "Survival Time (Weeks)",
           pval.coord = c(50,0.05), legend=c(0.75,0.9), legend.title="", title="Survival Probability (ASNS Quartile expression groups)", conf.int = FALSE,
           risk.table = TRUE) 

smoothCoxph(rppa_data$OS.Wks.SMK, rppa_data$status, rppa_data[[poi]], xlab="Protein Expression Levels")
```

## Multivariate Analysis 

```{r}
# Multivariate Analysis

rppa_data$Age.Dx <- ifelse(rppa_data$Age.Dx  > 60, "Greater", "Less")
rppa_data$Age.Dx  <- factor(rppa_data$Age.Dx , levels=c("Less", "Greater"))


covariate_names <- c(`Age.Dx:Greater` ="Age > 60", 
                     Cytogenetics.Risk.Group="Risk Level", 
                     `poi_median:Protein Expression < Median`="ASNS Low Expression", 
                     `poi_median:Protein Expression > Median`="ASNS High Expression",
                     `Cytogenetics.Risk.Group:Intermediate` = "ELN 2017 Intermediate Cytogenetics",
                     `Cytogenetics.Risk.Group:Unfavorable` = "ELN2017 Adverse Cytogenetics")
result <- rppa_data %>%
  analyse_multivariate(vars(OS.Wks.SMK, status==1),
                       covariates = vars(Age.Dx, Cytogenetics.Risk.Group, poi_median),
                       covariate_name_dict = covariate_names)
forest_plot(result,
            factor_labeller = covariate_names,
            endpoint_labeller = c(time="OS.Wks.SMK"),
            orderer = ~order(HR),
            labels_displayed = c("factor"),
            ggtheme = ggplot2::theme_bw(base_size = 8),
            relative_widths = c(1, 1.5, 1),
            HR_x_breaks = c(0.25, 0.5, 0.75, 1, 1.5, 2))

covariate_names <- c(`Age.Dx:Greater` ="Age > 60", 
                     Cytogenetics.Risk.Group="Risk Level", 
                     `poi_quartile:Q1`="ASNS Q1", 
                     `poi_quartile:Q2`="ASNS Q2",
                     `poi_quartile:Q3`="ASNS Q3",
                     `poi_quartile:Q4`="ASNS Q4",
                     `Cytogenetics.Risk.Group:Intermediate` = "ELN 2017 Intermediate Cytogenetics",
                     `Cytogenetics.Risk.Group:Unfavorable` = "ELN2017 Adverse Cytogenetics")

result <- rppa_data %>%
  analyse_multivariate(vars(OS.Wks.SMK, status==1),
                       covariates = vars(Age.Dx, Cytogenetics.Risk.Group, poi_quartile),
                       covariate_name_dict = covariate_names)
forest_plot(result,
            factor_labeller = covariate_names,
            endpoint_labeller = c(time="OS.Wks.SMK"),
            orderer = ~order(HR),
            labels_displayed = c("factor"),
            ggtheme = ggplot2::theme_bw(base_size = 8),
            relative_widths = c(1, 1.5, 1),
            HR_x_breaks = c(0.25, 0.5, 0.75, 1, 1.5, 2))

covariate_names <- c(`Age.Dx:Greater` ="Age > 60", 
                     Cytogenetics.Risk.Group="Risk Level", 
                     `poi_tertile:Q1`="ASNS Tertile Q1", 
                     `poi_tertile:Q2`="ASNS Tertile Q2",
                     `poi_tertile:Q3`="ASNS Tertile Q3",
                     `Cytogenetics.Risk.Group:Intermediate` = "ELN 2017 Intermediate Cytogenetics",
                     `Cytogenetics.Risk.Group:Unfavorable` = "ELN2017 Adverse Cytogenetics")

result <- rppa_data %>%
  analyse_multivariate(vars(OS.Wks.SMK, status==1),
                       covariates = vars(Age.Dx, Cytogenetics.Risk.Group, poi_tertile),
                       covariate_name_dict = covariate_names)
forest_plot(result,
            factor_labeller = covariate_names,
            endpoint_labeller = c(time="OS.Wks.SMK"),
            orderer = ~order(HR),
            labels_displayed = c("factor"),
            ggtheme = ggplot2::theme_bw(base_size = 8),
            relative_widths = c(1, 1.5, 1),
            HR_x_breaks = c(0.25, 0.5, 0.75, 1, 1.5, 2))
```